---
title: "Checkmate"
author: "Michel Lang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{checkmate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r,include=FALSE}
library(checkmate)
```

Ever used an R function that produced a not-very-helpful error message, just to discover after minutes of debugging that you simply passed a wrong argument?

Blaming the laziness of the package author for not doing such standard checks (in a dynamically typed language such as R) is at least partially unfair, as R makes theses types of checks cumbersome and annoying. Well, that's how it was in the past.

Enter checkmate.

Virtually **every standard type of user error** when passing arguments into function can be caught with a simple, readable line which produces an **informative error message** in case.
A substantial part of the package was written in C to **minimize any worries about execution time overhead**.

## Intro
As a motivational example, consider you have a function to calculate the faculty of a natural number and the user may choose between using either the stirling approximation or R's `factorial` function (which internally uses the gamma function).
Thus, you have two arguments, `n` and `method`.
Argument `n` must obviously be a positive natural number and `method` must be either `"stirling"` or `"factorial"`.
Here is a version of all the hoops you need to jump through to ensure that these simple requirements are met:
```{r}
fact <- function(n, method = "stirling") {
  if (length(n) != 1)
    stop("Argument 'n' must have length 1")
  if (!is.numeric(n))
    stop("Argument 'n' must be numeric")
  if (is.na(n))
    stop("Argument 'n' may not be NA")
  if (is.double(n)) {
    if (is.nan(n))
      stop("Argument 'n' may not be NaN")
    if (is.infinite(n))
      stop("Argument 'n' must be finite")
    if (abs(n - round(n, 0)) > sqrt(.Machine$double.eps))
      stop("Argument 'n' must be an integerish value")
    n <- as.integer(n)
  }
  if (n < 0)
    stop("Argument 'n' must be >= 0")
  if (length(method) != 1)
    stop("Argument 'method' must have length 1")
  if (!is.character(method) || !method %in% c("stirling", "factorial"))
    stop("Argument 'method' must be either 'stirling' or 'factorial'")

  if (method == "factorial")
    factorial(n)
  else
    sqrt(2 * pi * n) * (n / exp(1))^n
}
```
And for comparison, here is the same function using checkmate:
```{r}
fact <- function(n, method = "stirling") {
  assertCount(n)
  assertChoice(method, c("stirling", "factorial"))

  if (method == "factorial")
    factorial(n)
  else
    sqrt(2 * pi * n) * (n / exp(1))^n
}
```

## Function overview
All functions come in four flavors, depending on the prefix.
If prefixed with `assert`, an error is thrown if the check fails. Otherwise, `TRUE` is returned invisibly.
The family of functions prefixed with `test` always return the check result as logical value.
Functions starting with `check` return the error message as a string (or `TRUE` otherwise) and can be used if you need more control and, e.g., want to grep on the returned error message.
`expect` is the last family of functions and is intended to be used with the [testthat package](http://cran.r-project.org/package=testthat).
All performed checks are logged into the `testthat` reported.

### Scalars

* [checkFlag](http://www.rdocumentation.org/packages/checkmate/functions/checkFlag)
* [checkCount](http://www.rdocumentation.org/packages/checkmate/functions/checkCount)
* [checkInt](http://www.rdocumentation.org/packages/checkmate/functions/checkInt)
* [checkNumber](http://www.rdocumentation.org/packages/checkmate/functions/checkNumber)
* [checkString](http://www.rdocumentation.org/packages/checkmate/functions/checkString)
* [checkScalar](http://www.rdocumentation.org/packages/checkmate/functions/checkScalar)
* [checkScalarNA](http://www.rdocumentation.org/packages/checkmate/functions/checkScalarNA)
* [checkPercentage](http://www.rdocumentation.org/packages/checkmate/functions/checkPercentage)


### Vectors

* [checkLogical](http://www.rdocumentation.org/packages/checkmate/functions/checkLogical)
* [checkNumeric](http://www.rdocumentation.org/packages/checkmate/functions/checkNumeric)
* [checkInteger](http://www.rdocumentation.org/packages/checkmate/functions/checkInteger)
* [checkIntegerish](http://www.rdocumentation.org/packages/checkmate/functions/checkIntegerish)
* [checkComplex](http://www.rdocumentation.org/packages/checkmate/functions/checkComplex)
* [checkCharacter](http://www.rdocumentation.org/packages/checkmate/functions/checkCharacter)
* [checkFactor](http://www.rdocumentation.org/packages/checkmate/functions/checkFactor)
* [checkList](http://www.rdocumentation.org/packages/checkmate/functions/checkList)
* [checkVector](http://www.rdocumentation.org/packages/checkmate/functions/checkVector)
* [checkAtomic](http://www.rdocumentation.org/packages/checkmate/functions/checkAtomic)
* [checkAtomicVector](http://www.rdocumentation.org/packages/checkmate/functions/checkAtomicVector)


### Attributes

* [checkClass](http://www.rdocumentation.org/packages/checkmate/functions/checkClass)
* [checkNames](http://www.rdocumentation.org/packages/checkmate/functions/checkNames)
* [checkNamed](http://www.rdocumentation.org/packages/checkmate/functions/checkNamed)


### Choices and Subsets

* [checkChoice](http://www.rdocumentation.org/packages/checkmate/functions/checkChoice)
* [checkSubset](http://www.rdocumentation.org/packages/checkmate/functions/checkSubset)
* [checkSetEqual](http://www.rdocumentation.org/packages/checkmate/functions/checkSetEqual)


### Matrices, Arrays and Data Frame

* [checkMatrix](http://www.rdocumentation.org/packages/checkmate/functions/checkMatrix)
* [checkArray](http://www.rdocumentation.org/packages/checkmate/functions/checkArray)
* [checkDataFrame](http://www.rdocumentation.org/packages/checkmate/functions/checkDataFrame)


### Safe Coercion to integer

* [asCount](http://www.rdocumentation.org/packages/checkmate/functions/asInteger)
* [asInt](http://www.rdocumentation.org/packages/checkmate/functions/asInteger)
* [asInteger](http://www.rdocumentation.org/packages/checkmate/functions/asInteger)


### Other builtin

* [checkNull](http://www.rdocumentation.org/packages/checkmate/functions/checkNull)
* [checkEnvironment](http://www.rdocumentation.org/packages/checkmate/functions/checkEnvironment)
* [checkFunction](http://www.rdocumentation.org/packages/checkmate/functions/checkFunction)


### File IO:

* [checkFile](http://www.rdocumentation.org/packages/checkmate/functions/checkFile)
* [checkDirectory](http://www.rdocumentation.org/packages/checkmate/functions/checkDirectory)
* [checkPathForOutput](http://www.rdocumentation.org/packages/checkmate/functions/checkPathForOutput)


## In case you miss flexibility

You can use [assert](http://www.rdocumentation.org/packages/checkmate/functions/assert) to perform multiple checks at once and throw an assertion if all checks fail.


## Argument Checks for the Lazy

The following functions allow a special syntax to define argument checks using a special pattern.
E.g., `qassert(x, "I+")` asserts that `x` is an integer vector with at least one element and no missing values.
This provide a completely alternative mini-language (or style) how to perform argument checks.
You choose what you like best.

* [qassert](http://www.rdocumentation.org/packages/checkmate/functions/qassert)
* [qassertr](http://www.rdocumentation.org/packages/checkmate/functions/qassert)

## checkmate as testthat extension
To extend [testthat](http://cran.r-project.org/package=testthat), you need to IMPORT, DEPEND or SUGGEST on the `checkmate` package.
Here is a minimal example:
```{r,eval=FALSE}
# file: tests/test-all.R
library(testthat)
library(checkmate) # for testthat extensions
test_check("mypkg")
```
Now you are all set and can use more than 30 new expectations in your tests.
```{r,eval=FALSE}
test_that("checkmate is a sweet extension for testthat", {
  x = runif(100)
  expect_numeric(x, len = 100, any.missing = FALSE, lower = 0, upper = 1)
})
```

## Speed considerations
In comparison with tediously writing the checks yourself in R (c.f. factorial example at the beginning of the vignette), R is sometimes a tad faster while performing checks on scalars.
This seems odd at first, because checkmate is mostly written in C and should be comparably fast.
Yet many of the functions in the `base` package are not regular functions, but primitives.
While primitives jump directly into the C code, checkmate has to use the considerably slower `.Call` interface.
As a result, checkmate is sometimes slower while checking scalars or small vectors (but you have to invest some orders of magnitude more time to write these checks yourself).

For larger objects the tide has turned because checkmate avoids many unnecessary intermediate variables.
Here is a small unrepresentative benchmark.
But also note that this one here has been executed from inside `knitr` which is often the cause for outliers in the measured execution time.
Better run the benchmark yourself to get unbiased results.
```{r,dev="svg",fig.width=6,fig.height=4}
library(ggplot2)
library(microbenchmark)

# check x for being numeric of length 1000 with no missing nor NaN values
x = runif(1000)
r = function(x) stopifnot(is.numeric(x) && length(x) == 1000 && all(!is.na(x) & x >= 0 & x <= 1))
cm1 = function(x) assertNumeric(x, len = 1000, any.missing = FALSE, lower = 0, upper = 1)
cm2 = function(x) qassert(x, "N1000[0,1]")
mb = microbenchmark(r(x), cm1(x), cm2(x))
print(mb)
autoplot(mb)
```

```{r,dev="svg",fig.width=6,fig.height=4}
# check x for being a character vector with no missing values or empty strings
x = sample(letters, 10000, replace = TRUE)
r = function(x) stopifnot(is.character(x) && !any(is.na(x)) && all(nzchar(x)))
cm1 = function(x) assertCharacter(x, any.missing = FALSE, min.chars = 1)
cm2 = function(x) qassert(x, "S+[1,]")
mb = microbenchmark(r(x), cm1(x), cm2(x))
print(mb)
autoplot(mb)
```

```{r,dev="svg",fig.width=6,fig.height=4}
# check x for being a data frame with no missing values
N = 1000
x = data.frame(a = runif(N), b = sample(letters[1:5], N, replace = TRUE), c = sample(c(FALSE, TRUE), N, replace = TRUE))
r = function(x) is.data.frame(x) && !any(sapply(x, function(x) any(is.na(x))))
cm1 = function(x) testDataFrame(x, any.missing = FALSE)
cm2 = function(x) qtest(x, "D")
mb = microbenchmark(r(x), cm1(x), cm2(x))
print(mb)
autoplot(mb)

# checkmate tries to stop as early as possible
x$a[1] = NA
mb = microbenchmark(r(x), cm1(x), cm2(x))
print(mb)
autoplot(mb)
```
